---
title: "Cleaning script for AED data"
author: "Charles T T Edwards"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: no
vignette: >
  %\VignetteIndexEntry{data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.path = 'fig/clean-', fig.width = 6, tidy = TRUE, tidy.opts = list(blank = TRUE, width.cutoff = 120), message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>")
options(rmarkdown.html_vignette.check_title = FALSE)

suppressPackageStartupMessages(library(plyr))
suppressPackageStartupMessages(library(pander))
```

```{r, echo=TRUE}
# load data
load("GLOBAL_AED_1979_2016_raw.rda")
```

```{r}
# countries
countries <- unique(dat$country)

# cleaned data
dat_clean <- vector('list', length(countries))
names(dat_clean) <- countries

# summary table
tab <- data.frame(Country = countries, "no incomplete" = NA, "no duplicates" = NA, "no EXTR" = NA, "no E" = NA, "no averaged" = NA)
```

```{r}
# check for unknown countries
if (with(dat, any(is.na(country) | country == ""))) {
    warning("deleting unknown countries")
    dat <- subset(dat, !is.na(country) & country != "")    
}
```

```{r}
# loop over countries
for (i in countries) {

    dfr.clean <- subset(dat, country %in% i)

    # SIMPLE CHECKS FOR DATA QUALITY
    
    # remove incomplete data records
    dfr       <- dfr.clean
    dfr.clean <- subset(dfr, zone != "")
    dfr.clean <- subset(dfr.clean, !is.na(density))
    dfr.clean <- subset(dfr.clean, !is.na(year))
    dfr.clean <- subset(dfr.clean, !is.na(zone))
    
    tab[match(i, countries), 2] <- nrow(dfr) - nrow(dfr.clean)
    
    # remove duplicates
    dfr           <- dfr.clean
    dfr$duplicate <- FALSE
    dfr$duplicate[duplicated(dfr[, c('zone', 'country', 'type', 'year', 'estimate', 'survey_area')])] <- TRUE

    dfr.clean <- subset(dfr, !duplicate)
    
    tab[match(i, countries), 3] <- nrow(dfr) - nrow(dfr.clean)
    
    # remove extrapolations
    dfr       <- dfr.clean
    dfr.clean <- subset(dfr, type != "EXTR")
    
    tab[match(i, countries), 4] <- nrow(dfr) - nrow(dfr.clean)
    
    # remove low reliability surveys
    dfr       <- dfr.clean
    dfr.clean <- subset(dfr, reliability != "E")
    
    tab[match(i, countries), 5] <- nrow(dfr) - nrow(dfr.clean)
    
    # COMBINE SURVEYS FROM THE SAME ZONE IN THE SAME YEAR
    dfr       <- dfr.clean
    dfr.clean <- ddply(dfr, .(year, zone), summarize, 
                       region = unique(region),
                       subregion = unique(subregion),
                       survey_area_shrinkage = unique(survey_area_shrinkage),
                       reliability = max(as.character(reliability)), 
                       n_per_estimate = length(estimate), 
                       estimate = sum(estimate), 
                       survey_area = sum(survey_area), 
                       density = estimate / survey_area)
    
    tab[match(i, countries), 6] <- nrow(dfr) - nrow(dfr.clean)
    
    # END OF DATA QUALITY CHECKS
    dfr <- dfr.clean
    
    # CHECK SURVEY AREA SIZES TO ENSURE THAT ONLY SURVEYS THAT COVER A SIMILAR
    # AREA (KM2) PER ZONE ARE USED TO ESTIMATE THE TREND IN DENSITY FOR THAT 
    # ZONE
    
    # 4. identify surveys that are 20% "different" from
    # the modal area size. If any surveys in a time series are different
    # and survey is flagged as "max" then use max survey area for all years
    dfr.tmp <- ddply(dfr, .(zone), summarize, region = region, year = year, survey_area = survey_area, 
                     max_survey_area = rep(max(survey_area), length(survey_area)),
                     diff_from_mode_D20 = !redData::trim(x = survey_area, delta = 0.2)[['trim']], 
                     modal_survey_area = rep(redData::trim(x = survey_area), length = length(survey_area)),
                     use_max = ifelse(any(diff_from_mode_D20) & survey_area_shrinkage, TRUE, FALSE))
    
    dfr <- merge(dfr, dfr.tmp, all.x = TRUE)
    
    # identify surveys that are 40% different and (for the revised density)
	# have not had the survey area replaced by the max and 
	# mark for deletion
    dfr.tmp <- ddply(dfr, .(zone), summarize, region = region, year = year, survey_area = survey_area,
                     diff_from_mode_D40 = !redData::trim(x = survey_area, delta = 0.4)[['trim']], 
                     delete_revised = ifelse(diff_from_mode_D40 & !use_max, TRUE, FALSE),
					 delete = ifelse(diff_from_mode_D40, TRUE, FALSE))
    
    dfr <- merge(dfr, dfr.tmp, all.x = TRUE)
    
    # revised survey area and density
    dfr$revised_survey_area <- dfr$survey_area
    dfr$revised_survey_area[dfr$use_max] <- dfr$max_survey_area[dfr$use_max]
    
    dfr$revised_density <- dfr$estimate / dfr$revised_survey_area
    
    # mark for deletion
    dfr$raw_density <- dfr$density
    dfr$revised_density[dfr$delete_revised] <- NA
    dfr$density[dfr$delete] <- NA
    
    # count number surveys per zone
    dfr.tmp <- ddply(dfr, .(zone), summarize, n_surveys = length(density))
    dfr     <- merge(dfr, dfr.tmp, by = "zone", all = TRUE)

    # FORMAT AND CHECKING
    
    # check 
    stopifnot(nrow(dfr) == nrow(dfr.clean))
    
    # save cleaned data
    dat_clean[[i]] <- dfr
    
}
```

# Data quality checks

The following simple data quality checks were performed:

1) Delete incomplete records with: missing zone label, estimate, survey area, year or country
2) Delete records with identical country/year/type/estimate/survey values (i.e. duplicated)
3) Delete records with type = EXTR
4) Delete records with reliabilty = E
5) Sum survey data from same year and zone to give average density

```{r, echo = FALSE, results = "asis"}
pandoc.table(tab, style = "rmarkdown", split.tables = Inf)
```

# Data cleaning

Cleaning involved the preparation of data records for use in trend analysis. The first step involved identification of the modal survey area size. 

1) Identify surveys where the survey area size had shrunk over time in response to a contracting population range. 

Contraction of the survey range could lead to stabalisation of the density over time despite a declining population. Instances where this had occurred where identified using expert knowledge, and marked as `survey_area_shrinkage = TRUE`.

2) Identify modal area size

This was the most commonly observed survey area value or, when there was more than one area value observed with equal frequency, the modal survey area was the area that minimised the summed difference with other area values. If there were any ties, then both modal values were retained. 

3) Identify surveys with size >20\% different from modal sizes

Surveys with an area size of greater than 20\% difference to the nearest modal area size for that zone where identified; i.e. if there were more than one modal area size, then the minimum difference was used. These survey records were marked as `diff_from_mode_D20 = TRUE`.

4) Apply maximum survey area size to all records in a survey time series that have noticeable shrinkage.

If `survey_area_shrinkage == TRUE` and `any(diff_from_mode_D20) == TRUE` then the maximum survey area size was applied to all surveys in that series when calculating the density. These records were marked as `use_max = TRUE`

5) If any of the remaining surveys are >40\% from the mode the exclude that survey

If `use_max == FALSE` and `diff_from_mode_D40 = TRUE` then that particular survey was deleted (i.e. `delete = TRUE`).


```{r, results = "asis"}
# for each country print summary stats and plots

for (i in countries) {
    
    tab <- ddply(dat_clean[[i]], .(zone), summarize, 
                 mean_area = round(mean(survey_area),0), 
                 mean_revised_area = round(mean(revised_survey_area),0),
                 modal_area = unique(modal_survey_area), 
                 n_surveys = unique(n_surveys),
                 yrs_total = paste(range(year), collapse = "-"))

    print(paste("Summary data for", i))
    pandoc.table(tab, style = "rmarkdown", split.tables = Inf)
}
```

Melt list
```{r}
dat <- ldply(dat_clean, .id = "country")
```

Coerce factors
```{r}
dat$region    <- as.character(dat$region)
dat$subregion <- as.character(dat$subregion)
dat$country   <- as.character(dat$country)
dat$zone      <- as.character(dat$zone)
```

Delete sites with all zero
```{r}
dfr <- ddply(dat, .(zone), summarize, allzero = all(estimate == 0))

dfr <- subset(dfr, !allzero)

dat <- subset(dat, zone %in% dfr$zone)
```

```{r, echo = FALSE}
if (any(is.na(dat$density))) warning("NA values in density")
```

Write .csv file
```{r}
write.csv(dat, file = "GLOBAL_AED_1979_2016.csv", row.names = FALSE)
```

Save to package data
```{r, echo = TRUE}
# save clean data
AED <- dat; rm(dat)
suppressMessages(usethis::use_data(AED, overwrite = TRUE))
```



